\Rpackage{MSnbase}: labelled MS2 data pre-processing, visualisation and quantification.

### 2011-05-13 inst/doc/MSnbase-demo/MSnbase-demo.Rnw
updated and exported normalise method and vignette updates
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MSnbase@55561 bc3139a8-67e5-0310-9ffc-ced21a209358

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Section
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Importing raw data}\label{sec:msnexp}

@@

\Rpackage{MSnbase} imports raw MS data stored in \texttt{mzXML} 
format \citep{Pedrioli2004} with the function \Rfunction{readMzXMLData} function. 
See \Rfunction{?readMzXMLData} for more details.
More formats, notably \texttt{mzML} \citep{Martens2010}, will be added at a later stage. 
Either MS1 or MS2 spectra can be loaded at a time by setting the \texttt{msLevel} parameter accordingly.
In this document, we will use a slightly bigger dataset, \Robject{itraqdata}, provided with 
\Rpackage{MSnbase}. It includes feature metadata, accessible with the \Rfunction{fData} method. 
The metadata includes identification data for the \Sexpr{length(itraqdata)} MS2 spectra.

@@

    The experiment meta data associated with an \Robject{MSnExp} experiment
    is of class \Robject{MIAPE}. It stores general information about the 
    experiment as well as MIAPE (Minimum Information About a Proteomics Experiment, 
    \citep{Taylor2007, Taylor2008}) information.
    This meta-data can be accessed with the \Rfunction{experimentData} method. 
    When available, a summary of MIAPE-MS data can be printed with the \Rfunction{msInfo} method.
    See \texttt{?MIAPE} for more details.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Section
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Spectra objects}\label{sec:spectra}

The raw data is composed of the \Sexpr{length(itraqdata)} MS spectra. 
The spectra are named individually 
(\Sexpr{paste(paste(head(featureNames(itraqdata)),collapse=", "),", ...",sep="")})
and stored in a \Robject{environment}. They can be accessed individually with 
\texttt{itraqdata[["X10"]]} or as a list with \texttt{spectra(itraqdata)}.
As we have loaded our experiment specifying \texttt{msLevel=2}, 
the spectra will all be of level 2 (or higher, if available).

@@

Attributes of individual spectra or of all spectra of an experiment can be 
accessed with their respective methods: \Rfunction{precursorCharge} for the 
precursor charge, \Rfunction{rtime} for the retention time, \Rfunction{mz} 
for the MZ values, \Rfunction{intensity} for the inensities, \ldots see the 
\Robject{Spectrum}, \Robject{Spectrum1} and \Robject{Spectrum2} manuals for more details.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Section
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Reporter ions}\label{sec:reporterions}

@@

Spectra can be plotted individually or as part of (subsetted) experiments 
with the \Robject{plot} method. Full spectra can be plotted 
(using \texttt{full=TRUE}), specific reporter ions of interest 
(by specifing with reporters with \texttt{reporters=iTRAQ4} for instance) 
or both (see figure \ref{fig:spectrum-plot}).

@@

\begin{figure}[!ht]
<<msnexp-plot,fig=T,echo=T,tikz=T,external=F,width=5,height=6>>=
plot(bsa,reporters=iTRAQ4,full=FALSE)
@ 
\caption{Experiment-wide raw MS2 spectra. The y-axes of the individual 
  spectra are automatically rescaled to the same range. }
\label{fig:msnexp-plot}
\end{figure}

@@

\begin{figure}[!ht]
<<spectrum-plot2,fig=T,echo=F,tikz=T,external=F,width=5,height=2.5>>=
grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
vplayout <- function(x, y)
  viewport(layout.pos.row = x, layout.pos.col = y)
p1 <- plot(itraqdata[["X55"]],full=TRUE) + theme_gray(5) 
p2 <- plot(experiment[["X55"]],full=TRUE) + theme_gray(5) 
print(p1,vp=vplayout(1,1))
print(p2,vp=vplayout(1,2))
@ 
\caption{Same spectrum before (left) and after setting peaks <= 400 to 0.}
\label{fig:spectrum-plot2}
\end{figure}

@@

\begin{figure}[p]
  \begin{center}
<<preproc-plot,fig=T,echo=F,tikz=T,external=F,width=3,height=6>>=
grid.newpage()
pushViewport(viewport(layout = grid.layout(3, 1)))
print(p1, vp=vplayout(1,1))
print(p2, vp=vplayout(2,1))
print(p3, vp=vplayout(3,1))
@ 
\caption{This figure illustrated the effect or the \Rfunction{removePeaks} 
  and \Rfunction{clean} methods. The left-most spectrum displays two peaks, 
  of max height 3 and 7 respectively. The middle spectrum shows the result 
  of calling \Rfunction{removePeaks} with argument \Robject{t=3}, which sets 
  all data points of the first peak, whose maximum height is smaller or equal 
  to \Robject{t} to 0. The second peak is unaffected. Calling \Rfunction{clean} 
  after \Rfunction{removePeaks} effectly deletes successive 0 intensities from 
  the spectrum, as shown on the right plot. }
\label{fig:preproc}
  \end{center}
\end{figure}

Another useful manipulation method is \Rfunction{trimMz}, that takes as 
parameters and \Robject{MSnExp} (or a \Robject{Spectrum}) and a numeric 
\Robject{mzlim}. MZ values smaller then \texttt{min(mzlim)} 
or greater then \texttt{max(mzmax)} are discarded. This method is 
particularely useful when one wants to concentrate on a specific MZ range, 
as for reporter ions quantification, and generally results in substantial 
reduction of data size. Compare the size of the full trimmed experiment to 
the original \Sexpr{sz} Mb.
\label{trimMz-example}

@@

Quantitation is performed on fixed peaks in the spectra, that are specified 
with an \Robject{ReporterIon} object. A specific peak is defined by it's expected 
\Robject{mz} value and is searched for within \Robject{mz} $\pm$ \Robject{width}.  
If no data is found, \Robject{NA} is returned.

@@

The \Rfunction{quantify} method takes the following paramters: 
an \Robject{MSnExp} experiment, 
a character describing the quantification \Robject{method}, 
the \Robject{reporters} to be quantified 
and a \Robject{strict} logical defining whether data points 
ranging outside of \Robject{mz} $\pm$ \Robject{width} should be considered 
for quantitation. 
Additionally, a progress bar can be dispaying when setting 
the \Robject{verbose} parameter to \Robject{TRUE}.
Three quantification methods are implemented, as illustrated 
on figure \ref{fig:quant-methods}: 
\Robject{trapezoidation} returns the area under the peak of interest, 
\Robject{max} returns the apex of the peak and 
\Robject{sum} returns the sum of all intensities of the peak. 
See \texttt{?quantify} for more details.

@@

\begin{figure}[p]
<<quantitation-plot,fig=T,echo=F,tikz=T,external=F,width=5,height=5>>=
grid.newpage()
pushViewport(viewport(layout = grid.layout(2, 2)))
vplayout <- function(x, y)
  viewport(layout.pos.row = x, layout.pos.col = y)
print(p + opts(title="Quantitation using 'sum'") + geom_point(size=3,alpha=I(1/3),colour="red"), vp=vplayout(1,1))
print(p + opts(title="Quantitation using 'max'") + geom_point(aes(x=114.1,y=3),alpha=I(1/3),colour="red",size=3), vp=vplayout(1,2))
print(p + opts(title="Trapezoidation and strict=FALSE") + geom_polygon(alpha=I(1/5),fill="red"),vp=vplayout(2,1))
print(p + opts(title="Trapezoidation and strict=TRUE") + 
      geom_polygon(aes(x=c(114.05,114.05,114.1,114.15,114.15),y=c(0,1,3,1,0)),fill="red",alpha=I(1/5)),vp=vplayout(2,2))
@ 
\caption{The different quantitation methods are illustrated above. 
  Quantitation using \Robject{sum} sums all the data points in the peaks to 
  produce, for ths example, \Sexpr{quantify(sp,iTRAQ4[1],method="sum")[[1]]}, 
  whereas method \Robject{max} only uses the peak's maximum intensity, 
  \Sexpr{quantify(sp,iTRAQ4[1],method="max")[[1]]}. 
  \Robject{Trapezoidation} calculates the area under the peak taking 
  the full with into account (using \Robject{strict=FALSE} gives
  \Sexpr{round(quantify(ssp,iTRAQ4[1],method="trap",strict=FALSE)[[1]],3)}) 
  or only the width as defined by the reporter 
  (using \Robject{strict=TRUE} gives
  \Sexpr{round(quantify(ssp,iTRAQ4[1],method="trap",strict=TRUE)[[1]],3)}). }
\label{fig:quant-methods}
\end{figure}

@@

A \Robject{MSnSet} object is meant to be compatible with further downstream 
packages for data normalisation and statistical analysis. 
Below is an example of applying 
variance stabilization normalisation \citep{Huber2002} 
to the iTRAQ reporter intensities \citep{Karp2010}.

<<vsn,echo=TRUE,cache=FALSE>>=
library("vsn")
qnt.vsn <- vsn2(exprs(qnt))
@

Figure 3 of \cite{Karp2010} is reproduced with 
the dummy experiment and show on figure \ref{fig:cv-plot}.

<<cvdata,echo=FALSE,cache=FALSE>>=
sd1 <- apply(log2(exprs(qnt))+10,1,sd)
mn1 <- apply(log2(exprs(qnt))+10,1,mean)
cv1 <- sd1/mn1
sd2 <- apply(exprs(qnt.vsn)+10,1,sd)
mn2 <- apply(exprs(qnt.vsn)+10,1,mean)
cv2 <- sd2/mn2
library("zoo")
## rmed1 <- rollapply(cv1,7,function(x) median(x,na.rm=TRUE))
## rmed2 <- rollapply(cv2,7,function(x) median(x,na.rm=TRUE))
##
## Calling directly rollapply.zoo to make it zoo_1.6-4 compatible.
## The above requires zoo >= 1.7-0, which is as of 15 March 2011
## not yet available on CRAN (only on r-forge).
rmed1 <- zoo:::rollapply.zoo(cv1,7,function(x) median(x,na.rm=TRUE))
rmed2 <- zoo:::rollapply.zoo(cv2,7,function(x) median(x,na.rm=TRUE))
@

\begin{figure}[!ht]
<<cv-plot,fig=T,echo=F,tikz=T,external=F,width=5,height=5>>=
plot(order(mn1),cv1,col="steelblue",pch=19,
     ylab="CV",xlab="Rank mean",
     cex.axis=.7,cex.lab=.8)
points(order(mn2),cv2,col="red",pch=19)
grid()
lines(seq(1,70,by=(70/length(rmed1))),rmed1,col="steelblue",lwd=1)
lines(seq(1,70,by=(70/length(rmed2))),rmed2,col="red"      ,lwd=1)
## legend("topright",
##        legend=c("log2","vsn","running\nmedian"),
##        pch=c(19,19,NA),
##        col=c("steelblue","red","black"),
##        lwd=c(0,0,2),
##        bty="n",
##        cex=.7)
@ 
\caption{CV versus signal intensity comparison for log2 (blue) and vsn (red) transformed data. Lines indicate running CV medians.}
\label{fig:cv-plot}
\end{figure}

@@

\Rpackage{MSnbase} provides one function, \Rfunction{combineFeatures}, that allows 
to aggregate features stored in an \Robject{MSnSet} using build-in or user 
defined summary function and return a new \Robject{MSnSet} instance. 
The function has three main argument here described below. 
Additional details can be found in the function manual.

@@

<<makeGroups2,echo=TRUE,cache=FALSE>>=
gb <- fData(qnt)$ProteinAccession
table(gb)
@

The third argument, \Robject{fun}, 
defined how to combine the features. Predefined functions are readily available and can be 
specified as strings 
(\Robject{fun="mean"}, \Robject{fun="median"}, \Robject{fun="sum"}, 
\Robject{fun="weigthed.mean"} or \Robject{fun="medianpolish"} to compute respectively the mean, 
media, sum, weighted mean or median polish of the features to be aggregated). 
Alternatively, is is possible to supply userdefined functions with 
\Robject{fun=function(x) \{ ... \}}. We will use the \Robject{median} here.

@@

Impurities in the reporter reagents can bias the results and can be corrected when 
manufacturers provide correction coefficients. 
These generally come as percentages of each reporter ion that have masses differing 
by -2, -1, +1 and +2 Da from the nominal reporter ion mass due to isotopic variants.  
The \Rfunction{purityCorrect} method applies such correction to \Robject{MSnSet} 
instances. It also requires a square matric as second argument, \Robject{impurities}, 
that defines the repaltive percentage of reporter in the quantified each peak. 
See \texttt{?purityCorrect} for more details.

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Section
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Data structure and content}\label{sec:data}

\subsection{Importing experiments}\label{sec:msnexp}

@@

\Rpackage{MSnbase} imports raw MS data stored in \texttt{mzXML} 
format \citep{Pedrioli2004} with the function \Rfunction{readMzXMLData} function. 
See \Rfunction{?readMzXMLData} for more details.
More formats, notably \texttt{mzML} \citep{Martens2010}, will be added at a later stage. 
Either MS1 or MS2 spectra can be loaded at a time by setting the \texttt{msLevel} parameter accordingly.
In this document, we will use a slightly bigger data set, \Robject{itraqdata}, provided with 
\Rpackage{MSnbase}. It includes feature metadata, accessible with the \Rfunction{fData} method. 
The metadata includes identification data for the \Sexpr{length(itraqdata)} MS2 spectra.

@@

    The experiment meta data associated with an \Robject{MSnExp} experiment
    is of class \Robject{MIAPE}. It stores general information about the 
    experiment as well as MIAPE (Minimum Information About a Proteomics Experiment) 
    information \citep{Taylor2007, Taylor2008}.
    This meta-data can be accessed with the \Rfunction{experimentData} method. 
    When available, a summary of MIAPE-MS data can be printed with the \Rfunction{msInfo} method.
    See \texttt{?MIAPE} for more details.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Section
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Spectra objects}\label{sec:spectra}

The raw data is composed of the \Sexpr{length(itraqdata)} MS spectra. 
The spectra are named individually 
(\Sexpr{paste(paste(head(featureNames(itraqdata)),collapse=", "),", ...",sep="")})
and stored in a \Robject{environment}. They can be accessed individually with 
\texttt{itraqdata[["X1"]]} or \texttt{itraqdata[[1]]}, 
or as a list with \texttt{spectra(itraqdata)}.
As we have loaded our experiment specifying \texttt{msLevel=2}, 
the spectra will all be of level 2 (or higher, if available).

@@

Attributes of individual spectra or of all spectra of an experiment can be 
accessed with their respective methods: \Rfunction{precursorCharge} for the 
precursor charge, \Rfunction{rtime} for the retention time, \Rfunction{mz} 
for the MZ values, \Rfunction{intensity} for the intensities, \ldots see the 
\Robject{Spectrum}, \Robject{Spectrum1} and \Robject{Spectrum2} manuals for more details.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Section
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Reporter ions}\label{sec:reporterions}

@@

Spectra can be plotted individually or as part of (subset) experiments 
with the \Robject{plot} method. Full spectra can be plotted 
(using \texttt{full=TRUE}), specific reporter ions of interest 
(by specifying with reporters with \texttt{reporters=iTRAQ4} for instance) 
or both (see figure \ref{fig:spectrum-plot}).

@@

\begin{figure}[!ht]
<<msnexp-plot,fig=T,echo=T,tikz=T,external=F,width=5,height=6>>=
plot(bsa,reporters=iTRAQ4,full=FALSE)
@ 
\caption{Experiment-wide raw MS2 spectra. The y-axes of the individual 
  spectra are automatically rescaled to the same range. 
  See section \ref{sec:norm} to rescale peaks identically. }
\label{fig:msnexp-plot}
\end{figure}

@@

<<prepare-spectrum-plot2,echo=FALSE>>=
p1 <- plot(itraqdata[["X55"]],full=TRUE) + theme_gray(5) 
p2 <- plot(experiment[["X55"]],full=TRUE) + theme_gray(5) 
@

\begin{figure}[!ht]
<<spectrum-plot2,fig=T,echo=F,tikz=T,external=F,width=5,height=2.5>>=
grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
vplayout <- function(x, y)
  viewport(layout.pos.row = x, layout.pos.col = y)
print(p1,vp=vplayout(1,1))
print(p2,vp=vplayout(1,2))
@ 
\caption{Same spectrum before (left) and after setting peaks <= 400 to 0.}
\label{fig:spectrum-plot2}
\end{figure}

@@

\begin{figure}[p]
  \begin{center}
<<preproc-plot,fig=T,echo=F,tikz=T,external=F,width=3,height=6>>=
grid.newpage()
pushViewport(viewport(layout = grid.layout(3, 1)))
print(p1, vp=vplayout(1,1))
print(p2, vp=vplayout(2,1))
print(p3, vp=vplayout(3,1))
@ 
\caption{This figure illustrated the effect or the \Rfunction{removePeaks} 
  and \Rfunction{clean} methods. The left-most spectrum displays two peaks, 
  of max height 3 and 7 respectively. The middle spectrum shows the result 
  of calling \Rfunction{removePeaks} with argument \Robject{t=3}, which sets 
  all data points of the first peak, whose maximum height is smaller or equal 
  to \Robject{t} to 0. The second peak is unaffected. Calling \Rfunction{clean} 
  after \Rfunction{removePeaks} effectively deletes successive 0 intensities from 
  the spectrum, as shown on the right plot. }
\label{fig:preproc}
  \end{center}
\end{figure}

Another useful manipulation method is \Rfunction{trimMz}, that takes as 
parameters and \Robject{MSnExp} (or a \Robject{Spectrum}) and a numeric 
\Robject{mzlim}. MZ values smaller then \texttt{min(mzlim)} 
or greater then \texttt{max(mzmax)} are discarded. This method is 
particularly useful when one wants to concentrate on a specific MZ range, 
as for reporter ions quantification, and generally results in substantial 
reduction of data size. Compare the size of the full trimmed experiment to 
the original \Sexpr{sz} Mb.
\label{trimMz-example}

@@

Quantitation is performed on fixed peaks in the spectra, that are specified 
with an \Robject{ReporterIons} object. A specific peak is defined by it's expected 
\Robject{mz} value and is searched for within \Robject{mz} $\pm$ \Robject{width}.  
If no data is found, \Robject{NA} is returned.

@@

The \Rfunction{quantify} method takes the following parameters: 
an \Robject{MSnExp} experiment, 
a character describing the quantification \Robject{method}, 
the \Robject{reporters} to be quantified 
and a \Robject{strict} logical defining whether data points 
ranging outside of \Robject{mz} $\pm$ \Robject{width} should be considered 
for quantitation. 
Additionally, a progress bar can be displaying when setting 
the \Robject{verbose} parameter to \Robject{TRUE}.
Three quantification methods are implemented, as illustrated 
on figure \ref{fig:quant-methods}: 
\Robject{trapezoidation} returns the area under the peak of interest, 
\Robject{max} returns the apex of the peak and 
\Robject{sum} returns the sum of all intensities of the peak. 
See \texttt{?quantify} for more details.

@@

\begin{figure}[p]
<<quantitation-plot,fig=T,echo=F,tikz=T,external=F,width=5,height=5>>=
grid.newpage()
pushViewport(viewport(layout = grid.layout(2, 2)))
vplayout <- function(x, y)
  viewport(layout.pos.row = x, layout.pos.col = y)
print(p + opts(title="Quantitation using 'sum'") + geom_point(size=3,alpha=I(1/3),colour="red"), vp=vplayout(1,1))
print(p + opts(title="Quantitation using 'max'") + geom_point(aes(x=114.1,y=3),alpha=I(1/3),colour="red",size=3), vp=vplayout(1,2))
print(p + opts(title="Trapezoidation and strict=FALSE") + geom_polygon(alpha=I(1/5),fill="red"),vp=vplayout(2,1))
print(p + opts(title="Trapezoidation and strict=TRUE") + 
      geom_polygon(aes(x=c(114.05,114.05,114.1,114.15,114.15),y=c(0,1,3,1,0)),fill="red",alpha=I(1/5)),vp=vplayout(2,2))
@ 
\caption{The different quantitation methods are illustrated above. 
  Quantitation using \Robject{sum} sums all the data points in the peaks to 
  produce, for this example, \Sexpr{quantify(ssp,iTRAQ4[1],method="sum")[[1]]}, 
  whereas method \Robject{max} only uses the peak's maximum intensity, 
  \Sexpr{quantify(ssp,iTRAQ4[1],method="max")[[1]]}. 
  \Robject{Trapezoidation} calculates the area under the peak taking 
  the full with into account (using \Robject{strict=FALSE} gives
  \Sexpr{round(quantify(ssp,iTRAQ4[1],method="trap",strict=FALSE)[[1]],3)}) 
  or only the width as defined by the reporter 
  (using \Robject{strict=TRUE} gives
  \Sexpr{round(quantify(ssp,iTRAQ4[1],method="trap",strict=TRUE)[[1]],3)}). }
\label{fig:quant-methods}
\end{figure}

@@

A \Robject{MSnSet} object is meant to be compatible with further downstream 
packages for data normalisation and statistical analysis. There is also a 
\Rfunction{normalise} method for expression sets. The method takes and instance 
of class \Robject{MSnSet} as first argument, and a character to describe the 
\Robject{method} to be used:

\begin{description}
\item{\Robject{quantiles}} Applies quantile normalisation \citep{Bolstad03} as 
  implemented in the \Rfunction{normalize.quantiles} function of the 
  \Rpackage{preprocessCore} package. 
\item{\Robject{quantiles.robust}} Applies robust quantile normalisation 
  \citep{Bolstad03} as implemented in the 
  \Rfunction{normalize.quantiles.robust} function of the 
  \Rpackage{preprocessCore} package. 
\item{\Robject{vsn}} Applies variance stabilisation normalization \citep{Huber2002} 
  as implemented in the \Rfunction{vsn2} function of the 
  \Rpackage{vsn} package. 
\item{\Robject{max}} Each feature's reporter intensity is divided by the 
  maximum of the reporter ions intensities.
\item{\Robject{sum}} Each feature's reporter intensity is divided by the 
  sum of the reporter ions intensities.
\end{description}

<<normalise,echo=TRUE,cache=FALSE>>=
qnt.max <- normalise(qnt,"max")
qnt.sum <- normalise(qnt,"sum")
qnt.quant <- normalise(qnt,"quantiles")
qnt.qrob <- normalise(qnt,"quantiles.robust")
qnt.vsn <- normalise(qnt,"vsn")
@

The effect of these are illustrated on figure \ref{fig:norm-plot} and
figure \ref{fig:cv-plot} reproduces figure 3 of \cite{Karp2010} that described 
the application of vsn on iTRAQ reporter data.

\begin{figure}[p]
<<norm-plot,fig=T,echo=F,tikz=T,external=F,width=5,height=7>>=
.plot <- function(x,ttl=NULL) {
  boxplot(exprs(x),
          main=ifelse(is.null(ttl),processingData(x)@processing[6],ttl),
          cex.main=.8,
          cex.lab=.5,
          cex.axis=.5,
          cex=.8)
  grid()
}
oldmar <- par()$mar
par(mfrow=c(3,2),mar=c(2.9,2.9,2.9,1))
.plot(qnt,ttl="Non-normalised data")
.plot(qnt.max)
.plot(qnt.sum)
.plot(qnt.quant)
.plot(qnt.qrob)
.plot(qnt.vsn)
@ 
\caption{ Comparison of the normalisation \Robject{MSnSet} methods. 
  Note that vsn also glog-transforms the intensities. }
\label{fig:norm-plot}
\end{figure}

<<cvdata,echo=FALSE,cache=FALSE>>=
sd1 <- apply(log2(exprs(qnt))+10,1,sd)
mn1 <- apply(log2(exprs(qnt))+10,1,mean)
cv1 <- sd1/mn1
sd2 <- apply(exprs(qnt.vsn)+10,1,sd)
mn2 <- apply(exprs(qnt.vsn)+10,1,mean)
cv2 <- sd2/mn2
dfr <- rbind(data.frame(rank=order(mn1),cv=cv1,norm="raw"),
             data.frame(rank=order(mn2),cv=cv2,norm="vsn"))
library("zoo")
## rmed1 <- rollapply(cv1,7,function(x) median(x,na.rm=TRUE))
## rmed2 <- rollapply(cv2,7,function(x) median(x,na.rm=TRUE))
##
## Calling directly rollapply.zoo to make it zoo_1.6-4 compatible.
## The above requires zoo >= 1.7-0, which is as of 15 March 2011
## not yet available on CRAN (only on r-forge).
rmed1 <- zoo:::rollapply.zoo(cv1,7,function(x) median(x,na.rm=TRUE))
rmed2 <- zoo:::rollapply.zoo(cv2,7,function(x) median(x,na.rm=TRUE))
dfr2 <- 
  rbind(data.frame(x=seq(1,70,by=(70/length(rmed1))),
                   y=rmed1,norm="raw"),
        data.frame(x=seq(1,70,by=(70/length(rmed1))),
                   y=rmed2,norm="vsn"))

p <- qplot(rank,cv,data=dfr,col=norm) + 
  geom_line(data=dfr2,aes(x=x,y=y,colour=norm)) +
  theme_gray(7)
@

\begin{center}
\begin{figure}[!ht]
<<cv-plot,fig=T,echo=F,tikz=T,external=F,width=5,height=4>>=
print(p)
@ 
\caption{CV versus signal intensity comparison for log2 and vsn
  transformed data. Lines indicate running CV medians.}
\label{fig:cv-plot}
\end{figure}
\end{center}

Note that it is also possible to normalise individual spectra or whole experiments with 
the \Rfunction{normalise} method using the \Robject{max} method. 
This will rescale all peaks between 0 and 1. 
To visualise the relative reporter peaks, one should this first trim the spectra 
using method \Rfunction{trimMz} as illustrated in section \ref{sec:processing}, 
then normalise the \Robject{MSnExp} with \Rfunction{normalise} using 
\Robject{method="max"} as illustrated above and plot the data using \Rfunction{plot} 
(figure \ref{fig:msnexp-norm-plot}).

<<prepare-msnset-nrom-plot,cache=FALSE,echo=FALSE>>=
p <- plot(normalise(experiment[bsasel],"max"),reporters=iTRAQ4,full=FALSE)
p <- p + theme_gray(7)
@

\begin{figure}[p]
<<msnexp-norm-plot,fig=T,echo=F,tikz=T,external=F,width=5,height=6>>=
print(p)
@ 
\caption{Experiment-wide normalised MS2 spectra. The y-axes of the individual 
  spectra is now rescaled between 0 and 1 (highest peak), as opposed to 
  figure \ref{fig:msnexp-plot}. }
\label{fig:msnexp-norm-plot}
\end{figure}

@@

\Rpackage{MSnbase} provides one function, \Rfunction{combineFeatures}, that allows 
to aggregate features stored in an \Robject{MSnSet} using build-in or user 
defined summary function and return a new \Robject{MSnSet} instance. 
The three main arguments are described below. 
Additional details can be found in the method documentation.

@@

<<makeGroups2,echo=TRUE,cache=FALSE>>=
gb <- fData(qnt)$ProteinAccession
table(gb)
length(unique(gb))
@

The third argument, \Robject{fun}, 
defined how to combine the features. Predefined functions are readily available and can be 
specified as strings 
(\Robject{fun="mean"}, \Robject{fun="median"}, \Robject{fun="sum"}, 
\Robject{fun="weighted.mean"} or \Robject{fun="medianpolish"} to compute respectively the mean, 
media, sum, weighted mean or median polish of the features to be aggregated). 
Alternatively, is is possible to supply user defined functions with 
\Robject{fun=function(x) \{ ... \}}. We will use the \Robject{median} here.

@@

Impurities in the reporter reagents can bias the results and can be corrected when 
manufacturers provide correction coefficients. 
These generally come as percentages of each reporter ion that have masses differing 
by -2, -1, +1 and +2 Da from the nominal reporter ion mass due to isotopic variants.  
The \Rfunction{purityCorrect} method applies such correction to \Robject{MSnSet} 
instances. It also requires a square matrix as second argument, \Robject{impurities}, 
that defines the relative percentage of reporter in the quantified each peak. 
See \texttt{?purityCorrect} for more details.

@@

%% \section{Conclusion}\ref{sec:ccl}
